---
title: "CA Oil Spills 2008"
description: |
  Choropleth and interactive maps of oil spill incidents across CA in 2008.
author:
  - name: Sachiko Lamen
    url: {}
date: 2022-03-17
output:
  distill::distill_article:
    self_contained: false
    code_folding: TRUE
---

---
title: "CA Oil Spills 2008"
author: "Sachiko Lamen"
date: "3/1/2022"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include= FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(here)
library(broom)
library(sf) 
library(tmap) 
library(janitor)
library(spatstat)
library(maptools)
library(raster)
library(tidyverse)
```


### Overview:
This code creates maps depicting location and number of oil spills in each county in California in 2008. A G-function analysis was carried out to asses whether clustering is present in the data. See code for annotations. 

```{r, include = FALSE}
# Read in data
oil_spill_sf <- read_sf(here("data", "willamette", "oil_spill_shape", "Oil_Spill_Incident_Tracking_[ds394].shp")) %>%
  select(date = DATEOFINCI, county = LOCALECOUN)

ca_sf <- read_sf(here("data", "willamette", "ca_counties", "CA_Counties_TIGER2016.shp")) %>%
  clean_names() %>%
  select(county_name = name, land_area = aland)

# View CRS info
oil_spill_sf %>% st_crs()
ca_sf %>% st_crs()
# Looks like they are the same! Yay!
```

### Exploratory Interactive Map
First, lets see what our data looks like using a basic interactive tmap:

```{r}
# Create basic interactive exploratory plot using tmap
tmap_mode(mode = 'view')
tm_shape(ca_sf) +
  tm_borders(col = 'black') +
  tm_shape(oil_spill_sf) +
  tm_dots()
```

**Figure 1.** Black points represent oil spills in California in 2008.

### Choropleth Map: Oil Spill Incidents per County
Make a map that colors counties based on the relative number of oil spills that occurred in each county in 2008.

```{r}
# Join oil spill and California counties data sets/geometry so we have themn all in one place
ca_oil_sf <- ca_sf %>%
  st_join(oil_spill_sf)

# Count number of oil spills per county (exclude counties that have NA values)
oil_counts_sf <- ca_oil_sf %>%
  group_by(county_name) %>%
  summarize(n_records = sum(!is.na(county_name))) 

# Create a choropleth map using geom_sf()
ggplot(data = oil_counts_sf) +
  geom_sf(aes(fill = n_records), color = 'white', size = 0.1) +
  scale_fill_gradientn(colors = c('lightgrey', 'darkorchid', 'navyblue')) +
  theme_void() +
  labs(fill = "Number of Oil Spills",
       title = "CA Oil Spills Per County 2008")
```


**Figure 2.** Oil spills per county in California in 2008. Darker shades of purple represent larger numbers of oil spills.

```{r}
# Just for fun, make a cool interactive choropleth map using tmap
tmap_mode(mode = 'view')
tm_shape(oil_counts_sf) +
  tm_borders(col = 'black') +
  tm_fill('n_records', palette = 'BuPu', title = "Number of Oil Spills") 
```

**Figure 3.** *Same map as above, but interactive:* Oil spills per county in California in 2008. Darker shades of purple represent larger numbers of oil spills.

### Nearest Neighbor G-Function Analysis
Lets say we want to find out whether or not these oil spill incidents are clustered or random. We can find out using a G-function analysis.

```{r}
oil_sp <- as(oil_spill_sf, 'Spatial') # Convert to object 'Spatial'
oil_ppp <- as(oil_sp, 'ppp') # Convert to spatial point pattern

ca_sp <- as(ca_sf, 'Spatial') # Convert to object 'Spatial'
ca_win <- as(ca_sp, 'owin') # this window will exclude marine oil spills, but there is still enough data to perform G Function analysis without these observations

# Combine as a point pattern object (points + window):
oil_full <- ppp(oil_ppp$x, oil_ppp$y, window = ca_win) 

# make vector containing values from 0 - 10,000 that will be used to calculate G(r)
r_vec <- seq(0, 10000, by = 100) 
```

```{r, message = FALSE, warning = FALSE}
gfunction <- envelope(oil_full, fun = Gest, r = r_vec, nsim = 20, nrank = 2) # Calculate the actual and theoretical G(r) values, using 20 simulations of CRS for the "theoretical" outcome (the processing time is very low so fewer simulations will run faster)

# Create dataframe to be used for G-function plot
gfunction_long <- gfunction %>%
  as.data.frame() %>%
  pivot_longer(cols = obs:hi, names_to = 'model', values_to = 'g_val')
```


```{r}
# Create basic graph of G-function analysis
ggplot(data = gfunction_long, aes(x = r, y = g_val, group = model)) +
  geom_line(aes(color = model)) +
  labs(title = "G-function Plot", group = "Model")
```

**Figure 3.**  G Function Analysis. 

This confirms clustering - our data has a greater proportion of events with nearest neighbor at *smaller distances* compared to a theoretical CSR scenario.


### Citation:
CA DFW Oil Spill Incident Tracking <https://gis.data.ca.gov/datasets/7464e3d6f4924b50ad06e5a553d71086_0/data>











