---
title: "Mono Lake"
description: |
  Graphical analysis of historical lake level measurments with superimposed policy milestones.
author:
  - name: Sachiko Lamen
    url: 
date: 2022-02-21
output:
  distill::distill_article:
    self_contained: false
    code_folding: TRUE
---

```{r setup, include = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(plotly)
library(ggnewscale)
library("png")
library("grid")
```

## **Overview**
The following code is used to create a visually appealing representation of historic Mono Lake water levels and highlights important environmental and policy milestones (see below). See code for annotations.

```{r, message = FALSE, warning = FALSE}
# Read in and wrangle the data: remove metadata rows, rename columns
mono <- read_csv(here("data", "mono_lake_annual_levels.csv"), col_names = FALSE, skip = 6) %>%
  rename(year = X1, lake_level = X2, stable_level = X3, volume = X4, surface_area = X5, vert_change = X6)

# import a plot background image
lake_image <- readPNG(here("data", "Mono_Lake_from_Mount_Dana_(1).png"))
```

```{r}
# Create graph of lake level change from 1850 - 2017, with labels for important milestones
ggplot(data = mono, aes(x = year, y = lake_level)) +
  scale_fill_continuous(guide = "none") + 
  annotation_custom(rasterGrob(lake_image, 
                               width = unit(1,"npc"), 
                               height = unit(1,"npc")), 
                    -Inf, Inf, -Inf, Inf) + # This section replaces background with image of Mono Lake
  geom_ribbon(data = mono %>% filter(lake_level <= 6392),
              aes(ymin = lake_level, ymax = 6392), 
              fill = "firebrick1", alpha = 0.5) + # This section creates translucent red fill below the graphed line
  geom_line() +
  geom_hline(yintercept = 6392, linetype = "dashed", color = "red") + # add horizontal dashed line indicating stable lake levels
  geom_hline(yintercept = 6377, linetype = "dashed", color = "orange") + # add horizontal dashed line indicating when level at which land bridges emerge
  geom_hline(yintercept = 6360, linetype = "dashed", color = "grey") + # add horizongal dashed line indicating lake level when salinity reaches critical concentration
  annotate(
    geom = "curve", x = 1935, y = 6412, xend = 1941, yend = 6417, 
    curvature = .3, arrow = arrow(length = unit(2, "mm")),
    color = "white"
  ) + # this section adds an arrow for label_1
  annotate(geom = "text", x = 1934, y = 6409.5, label = "DWP begins 
  diverting water", size = 3, hjust = "right", color = "white") + # this section adds text for label_1
  annotate(geom = "point", x = 1941, y = 6417, color = "yellow", size = 2, alpha = 0.5) + # this section adds a marker point for label_1
  annotate(
    geom = "curve", x = 1963, y = 6370, xend = 1979, yend = 6373.4, 
    curvature = .1, arrow = arrow(length = unit(2, "mm")),
    color = "white"
  ) + # this section adds an arrow for label_2
  annotate(geom = "text", x = 1962, y = 6370, label = "MLC sues DWP", size = 3, hjust = "right", color = "white") + # this section adds text for label_2
  annotate(geom = "point", x = 1979, y = 6373.4, color = "yellow", size = 2, alpha = 0.5) + # this section adds a marker point for label_2
   annotate(
    geom = "curve", x = 1979, y = 6394.1, xend = 1983, yend = 6378.6, 
    curvature = .3, arrow = arrow(length = unit(2, "mm")),
    color = "white"
  ) + # this section adds an arrow for label_3
  annotate(geom = "text", x = 1971, y = 6395, label = "Supreme Court rules
           in favor of MLC", size = 3, hjust = "left", color = "white") + # this section adds text for label_3
  annotate(geom = "point", x = 1983, y = 6378.6, color = "yellow", size = 2, alpha = 0.5) + # this section adds a marker point for label_3
   annotate(
    geom = "curve", x = 1985, y = 6365, xend = 1994, yend = 6374.6, 
    curvature = .3, arrow = arrow(length = unit(2, "mm")),
    color = "white"
  ) + # this section adds an arrow for label_4
  annotate(geom = "text", x = 1985, y = 6364.9, label = "Decision 1631", size = 3, hjust = "right", color = "white") + # this section adds text for label_4
  annotate(geom = "point", x = 1994, y = 6374.6, color = "yellow", size = 2, alpha = 0.5) + # this section adds a marker point for label_4
  annotate(geom = "text", x = 1850, y = 6394, label = "Stable Level", size = 3.5, hjust = "left", color ="red") + # this section adds text label for the horizontal dashed line indicating stable level
  annotate(
    geom = "curve", x = 2005, y = 6370, xend = 1998, yend = 6384.3, 
    curvature = .1, arrow = arrow(length = unit(2, "mm")),
    color = "white"
  ) + # this section adds an arrow for label_5
  annotate(geom = "point", x = 1998, y = 6384.3, colour = "yellow", size = 2, alpha = 0.5) + # this section adds a marker for label_5
  annotate(geom = "text", x = 2011, y = 6366.9, label = "Restoration orders
issued by state", size = 3, hjust = "center", color = "white") + # this section adds text for label_5
  annotate(geom = "text", x = 1850, y = 6379, label = "Land bridges emerge", size = 3.5, hjust = "left", color ="orange") + # this section adds text label to horizontal dashed line for land bridges
  annotate(geom = "text", x = 1850, y = 6362, label = "Salinity 120 g/L", size = 3.5, hjust = "left", color ="grey") + # this section adds a text label to the horizontal dashed line for salinity concentrations
    labs(x = "Year",
         y = "Lake Level (ft above sea level)",
         title = "Mono Lake level changes (1850 - 2017)") # this section creates axis titles and main title
  

```

**Figure 1. Mono Lake Historical Lake Level (1850 - 2017)** Annual observations of Mono Lake level (ft above sea level) are tracked by the black line. The red dashed line represents the stable level (6392 ft) designated by the State Water Board that the DWP is required to restore Mono Lake to. The orange dashed line represents the lake level at which land bridges are exposed (6379 ft), making critical shorebird nesting sites vulnerable to terrestrial predators. The grey dashed line represents the lake level where the salinity of the water reaches 120 g/L (6360 ft); critical food resources for aquatic birds would be negatively affected at this level of salinity and acutely impacted at concentrations higher than 150 g/L (<6350 ft). (**DWP**: Los Angeles Department of Water and Power, **MLC**: Mono Lake Committee, **Decision 1631**: mandated that Mono Lake be allowed to rise to 6392 ft above sea level and set water export rules for DWP depending on current lake level, **Restoration Orders 98-05 and 98-07**: mandated raising Mono Lake to 6392 ft and improving existing freshwater ponds for migrating birds)

